### 2022-10-12 dissertation/chapters/02-artificial-neural-networks.tex
✨ (dissertation): Chapter 2 WIP

\begin{quote}
    \textit{There are an estimated 100 billion neurons in the human brain. That
        means that there are approximately as many neurons in the human brain as
        there are stars in our galaxy. Every one of these neurons is highly
        interconnected with other neurons, yielding a synapse count of 100
        trillion. These neural structures are the result of millions of years
        of evolution. It is with these neural structures that we are able to think,
    learn and dream.}
\end{quote}

The human brain is by far one of the most complicated biological organs in
nature. The human brain is what gives us the ability to think and learn. The
field of \ac{ML} has taken a lot of inspiration from the biological brain which
lead to the development of \acp{ANN} \cite{ref:rosenblatt:1958}. \acp{ANN}
are used throughout this dissertation as the model to be trained using \acp{HH}.
Understanding how \acp{ANN} work is an important component of this dissertation.

The purpose of this chapter is to provide the necessary background information
needed on \acp{ANN} and is thus structured into five main parts:

@@

    \item
    \textbf{Section \ref{sec:anns:an}} introduces the \ac{AN} and the various
    components that it is made up of. Brief discussions follow on
    input, \index{weights}weights, \index{net input
    signal}net input signal, \index{biases}biases,
    \index{activation functions}activation functions, and
    output.

    \item
    Discussions on the \ac{BN} and the \ac{AN} set the tone for the \ac{ANN}
    which is discussed in detail in \textbf{Section \ref{sec:anns:ann}}. Brief
    discussions follow on \ac{ANN} architecture, topology, and
    \acp{FFNN}.

    \item
    \textbf{Section \ref{sec:anns:training}} gives details on the training process,
    training sets, \index{supervised learning}supervised learning,
    \index{error function}error functions, performance measures
    and stopping conditions.

    \item
    Finally, a brief summary of the chapter is given in
    \textbf{Section \ref{sec:anns:summary}}.
\end{itemize}

\index{Biological Neuron}
\section{Biological Neuron}
\label{sec:anns:bn}

This section introduces the \ac{BN} and provides the necessary background
information in order to understand how it has inspired the development
of the \ac{AN}.

@@

\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.8\textwidth]{biological_neuron.eps}
    \caption[The \index{Biological Neuron}Biological Neuron]{An illustration of
    the \index{Biological Neuron}Biological Neuron.}
    \label{fig:biological_neuron}
\end{figure}

The main parts of the \ac{BN} to focus on include the cell body, the
\index{dendrites}\textit{dendrites} and the \index{axon}\textit{axon} because these
components are directly modeled in the \ac{AN}. \Acp{NN} are formed by
neurons that connect to each other. This is known as
\index{synaptogenesis}\textit{synaptogenesis} \cite{ref:huttenlocher:1997}.
Connections are made between the \index{axons}axons and
\index{dendrites}dendrites of various neurons. Such a connection is referred to
as a \index{synapse}\textit{synapse}. It is through \index{synapses}synapses
that neurons communicate with each other. Communication takes place by
electro-chemical pulse and is often referred to as an \textit{activation} or
\textit{action potential}.  Communication signals propagate from the
\index{dendrites}dendrites, through the cell body to the axon of a neuron,
provoking a signal in the post-synaptic neuron \cite{ref:engelbrecht:2007}.
Neurons like these are massively interconnected.  The greater the connection
between two neurons, the stronger the communication.  Kennedy
\cite{ref:kennedy:2016} defines stronger \index{Synapse}synapses as ones that
contribute more depolarisation to the neural membrane upon activation than
weaker ones. Stronger \index{Synapse}synapses have a higher probability of
generating an action potential in the post-synaptic neuron. During activation,
the pre-synaptic neuron release neurotransmitters that bind to the post-synaptic
neuron \cite{ref:khanacademy:synapse}. The frequent release of these molecules
cause the \index{synapse}synapse to grow. Connections that grow over time yield
stronger signals, while connections that are weak propagate low intensity
signals and vanish over time. This is how the human brain learns and forgets.
Every time we reinforce a memory in our brains, we ``train'' the neuron and the
connection becomes stronger. This is known as \index{synaptic
plasticity}\textit{synaptic plasticity} \cite{ref:huttenlocher:1997}. this
reinforcement of connections plays a pivotal role in the way we algorithmically
train \acp{ANN}.

\index{Artificial Neuron}
\section{Artificial Neuron}
\label{sec:anns:an}

This section introduces the \ac{AN}. It is shown how the \ac{AN} is modeled
after the \ac{BN} by providing a brief discussion on the various components that
make up the \ac{AN}. Each component is discussed in more detail, including input
modeling, weight modeling, \index{net input signal}net input signal modeling,
biases modeling and output modeling in the \ac{AN}.

\subsection{Components}
\label{sec:anns:components}

The \ac{AN} implements a nonlinear mapping from $\mathbb{R}^{I}$ to
$\mathbb{R}^{T}$, usually in the ranges $[0,1]$ or $[-1,1]$, depending on the
activation function used \cite{ref:engelbrecht:2007}. This mapping is given
below:

@@

\noindent
where $f_{AN}$ is the function mapping produced by the \ac{AN}, $I$
is the total number of dimensions of the input in real-number space $\mathbb{R}$,
and $T$ is the total number of dimensions of the target (desired output) in
real-number space.

An illustration of the \ac{AN} is given is Figure \ref{fig:artificial_neuron}.

\begin{figure}[htpb]
  \centering
  \includegraphics[width=\textwidth]{artificial_neuron.eps}
  \caption[The \index{Artificial Neuron}Artificial Neuron]{An illustration of
  the \index{Artificial Neuron}Artificial Neuron.}
  \label{fig:artificial_neuron}
\end{figure}

From Figure~\ref{fig:artificial_neuron}, one can see that the
\ac{AN} is made up of the following components:

\begin{itemize}
    \item
    \textbf{Input:} The input models the activations received from the pre-synaptic neuron
    or from some environment sensor. Input is represented by the input vector
    $\vec{x}$.

    \item
    \textbf{Weights:} The weights model the \index{synapse}synapses and connection
    strengths. Weights are represented by the weight vector, $\vec{v}$.

    \item
    \index{net input signal}\textbf{Net input signal:} The \index{net input signal}net input signal models
    the net resulting activations from all connected pre-synaptic neurons or
    environment sensors. The \index{net input signal}net input signal is
    represented by $net$.

    \item
    \textbf{Biases:} The biases model a mechanism introduced to influence the strength of the
    activation (output signal) of the \index{AN}\ac{AN}. The bias is
    represented by $\theta$.

    \item
    \index{activation function}\textbf{Activation function:} Models the action potential of the
    \ac{BN}\index{BN} based on the net input signal. The \index{activation
    function}activation function is
    represented by $f$.

    \item
    \textbf{Output:} The output models the activation by the post-synaptic neuron. Output is
    represented by the output vector $\vec{y}$.
\end{itemize}

Each of the above mentioned components are discussed in detail in the following
sections.

\subsection{Input}
\label{sec:anns:an:input}

Input signals are $I$-dimensional vectors of numerical values that are obtained
either through some environment sensor or from other \acp{AN}. Input signals
are often referred to as \index{features}\textit{features} or \index{independent
variables}\textit{independent variables} \cite{ref:francis:2001}. In this
dissertation, a single input vector is referred to as a pattern. Input data must first
be pre-processed before it is presented to the \ac{AN}. Input pre-processing
techniques are discussed next.

\subsubsection{Input Pre-processing}
\label{sec:anns:an:input:input_pre_processing}

Input pre-processing can help improve the training process, but also plays an
important role in determining the success of a practical application of the
\ac{ANN} \cite{ref:kuzniar:2017}. The primary purpose of input
pre-processing is to modify the input data so that it can better match predicted
output.

There are many techniques to consider when pre-processing input data. From
scaling to the correct ranges to dealing with incomplete, invalid or even
irrelevant data. For the purposes of this dissertation, only data encoding and
normalisation is considered. These topics are now discussed in more detail.

\index{Encoding}
\subsubsection{Encoding}
\label{sec:anns:an:input:encoding}

For qualitative data, such as class labels, the class labels have to be
converted to numerical representations. In general, class labels are encoded as
vectors \cite{ref:srinidhi:2018, ref:brownlee:2017:one-hot}. This dissertation makes
use of \index{One-Hot Encoding}\textit{one-hot encoding} of class labels which
yields a $K$-dimensional output vector $\vec{y}$ of $0$'s and a single $1$. Each
element $y_k$ of $\vec{y}$ uniquely refers to a class. An activation, represented by a value
of $1$ at a given index thus represents a given class. Harris and Harris
\cite{ref:harris:2010} describe \index{One-Hot Encoding}one-hot encoding as a
group of vector elements where the classes are represented
with a single high element (represented by $1$) and all the others low
(represented by $0$).

Once data has been encoded into the right form, normalisation is required. This
is discussed next.

\index{Normalisation}
\subsubsection{Normalisation}
\label{sec:anns:an:input:normalisation}

Input data must be normalised and scaled appropriately. In general, input data
is normalised to a range that is appropriate for the \index{Activation
Function}activation function being used. This yields input data in a range that
is comparable to the output produced. For the purpose of this dissertation, the
min-max scaler \cite{ref:al:2006} and standard score scaler \cite{ref:jain:2005}
are normalisation techniques that are considered. These are discussed next.

\index{Min-Max Scalar}
\subsubsection{Min-Max Scaler}
\label{sec:anns:an:input:min_max_scaler}

The \index{Min-Max Scaler}min-max scaler scales all values into the range
$[0,1]$. This is also called \index{Unity-based
Normalisation}\textit{unity-based normalisation}. The
\index{min-max scaler}min-max scaler is used as a pre-processing technique for
target values when the \index{sigmoid}\textit{sigmoid} \index{activation
function} activation function is used and is given below:

@@

\noindent where $x_{i,p}^{'}$ is the normalised form of $x_{i,p}$, the $i$-th
dimension of the input vector $\vec{x}_p$, $p \in \{1,2, \dots, P \}$, where $P$
is the total number of input patterns. $x_{i_{min}}$ and $x_{i_{max}}$ are
respectfully the minimum and maximum values of the $i$-th dimension for all
input vectors $\vec{x}_p, p \in \{1,2, \dots, P\}$.

\index{Standard Score Scalar}
\subsubsection{Standard Score Scaler}
\label{sec:anns:an:input:standard_score_scaler}

The \index{standard score scaler}standard score scaler, also known as the
\index{z-score scaler}\textit{z-score scaler}, scales features by subtracting
the mean and scaling to the unit variance of each dimension $i$ for all input
patterns $\vec{x}_p, p \in \{1,2, \dots, P\}$. The \index{standard score
scaler}standard score scaler is used as a pre-processing technique for target
values when the \index{hyperbolic tangent}\textit{hyperbolic tangent}
\index{activation function}activation function\index{activation function} is
used. The normalised form of $x_{i,p}^{'}$ as determined by the \index{standard
score scaler}standard score scaler is given in below:

@@

\noindent where $\mu_i$ and $\sigma^2_i$ are respectfully the mean and unit
variance of the $i$-th dimension of all input vectors $\vec{x}_p, p \in \{1,2,
\dots, P\}$.

\subsection{Weights}
\label{sec:anns:an:weights}

It was mentioned in Section~\ref{sec:anns:bn} that \index{synapses}synapses can
have strong or weak connections. The connection strength is modeled in the
\ac{AN} as weight vectors of numerical values associated with each dimension $i$
of the input data. Weights can either dampen or strengthen the input data by
some negative or positive numerical value. By changing the weight associated
with a feature, the influence of that particular feature on the predicted output
is changed. Finding the correct value for each weight, to yield optimal output,
is an optimisation problem \cite{ref:thierens:1993}. The initial placement of
these weight vectors in the search-space is very important. Research has shown
that weight initialisation plays an important role in the efficient training of
\acp{ANN} \cite{ref:thimm:1995}. Weight initialisation is discussed next,
followed by some common weight initialisation techniques used.

\index{Weight Initialisation}
\subsubsection{Weight Initialisation}
\label{sec:anns:an:weights:initialisation}

Weight initialisation is the process by which candidate solutions (represented
by the weights of the \ac{AN}) to the problem are placed in the search-space.
Weight initialisation influences the speed of convergence, the probability of
convergence and the generalisation capabilities of \acp{ANN}
\cite{ref:fernandez:2001}. Finding the optimal initialisation values for weights
is non-trivial and can be seen as another optimisation problem. Extensive
research on this topic has been done in the past \cite{ref:de:2016,
ref:erdogmus:2003, ref:yam:2000}.

\index{Weight Initialisation}Weight initialisation is dependent on the
\index{activation function}activation function used. If weights are initialised
to values that are too small, the vanishing gradients problem can occur
\cite{ref:hanin:2018}. If weights are initialised to values that are too big,
output of the activation function would not be in the active range. This leads
to unit saturation, and exploding gradients can occur
\cite{ref:hanin:2018, ref:yadav:2018}.

Research has presented many initialisation techniques \cite{ref:erdogmus:2003}.
Some techniques have been developed to address the problems that are mentioned
above \cite{ref:yadav:2018}. However, this dissertation will only focus on
\index{random uniform sampling}random uniform sampling, \index{Glorot uniform
sampling}\textit{Glorot uniform (Xavier)} sampling and \index{Glorot normal
sampling}\textit{Glorot normal} sampling \cite{ref:glorot:2010}. These
techniques are briefly discussed next.

\index{random uniform sampling}
\subsubsection{Random Uniform Sampling}
\label{sec:anns:an:weights:random_uniform_sampling}

\index{random uniform sampling}Random uniform sampling initialises weights
uniformly in the range $[\omega_{min}, \omega_{max}]$, written as $w_{i} \sim
\textit{U} (\omega_{min}, \omega_{max})$ where the $\omega_{min}$ and
$\omega_{max}$ are respectfully the lower and upper bounds of the uniform
distribution. Suggested parameter values are $(-1, 1)$ or $(-0.5, 0.5)$
\cite{ref:nguyen:1990}.

\index{Glorot uniform sampling}
\subsubsection{Glorot Uniform Sampling}
\label{sec:anns:an:weights:glorot_uniform_sampling}

@@

\index{Glorot Normal Sampling}
\subsubsection{Glorot Normal Sampling}
\label{sec:anns:an:weights:glorot_normal_sampling}

@@

\index{net input signal}
\subsection{Net Input Signal}
\label{sec:anns:an:net_input}

\acp{AN} accumulate the net resulting input signal from all input dimensions
into a value called the \index{net input signal}\textit{net input signal},
expressed as $net$. This signal is passed to the \index{Activation
Function}activation function, which provokes an artificial action potential in
the \ac{AN}.

The most common way by which the \index{net input signal}net input signal is
calculated is by means of \acp{SU} \cite{ref:engelbrecht:2007}, which
compute the \index{net input signal}net input signal as the weighted sum of all
input signals as shown below:

@@

\noindent where $x_{i}$ is the $i$-th
dimension of the input vector $\vec{x}$, and $v_{i}$ is the $i$-th dimension of
the weight vector $\vec{v}$, associated with the given input dimension.

\subsection{Biases}
\label{sec:anns:an:biases}

A bias/threshold term $\theta$ is introduced to help translate the output of the
\index{activation function}activation function \cite{ref:benitez:1997}. The
value of $\theta$ must be learned during the training process along with the
weights of the \ac{ANN}. In order to simplify equations, augmented vectors are
used in which input and hidden layers have an additional neuron/unit, called the
\textit{bias unit} \cite{ref:engelbrecht:2007}. Augumented vectors that include
the bias unit to the net input signal are presented next.

\subsubsection{Augmented Net Input Signal}
\label{sec:anns:an:biases:augmented_net_input_signal}

Input layer $x_{I+1}$ has a net input signal $net^{'} = net - \theta$. The value
for $\theta$ is learned along with the weights such that $\theta =
x_{i+1}v_{i+1}$. If a constant value $x_{i+1} = -1$ is used, the equation for
\acp{SU} as given in Equation~\eqref{eq:summation_units} is then completed by
including $\theta$ as shown below:

@@

\noindent where $\textit{net}^{'}$ is the augmented net input signal that now
includes the bias term $\theta$.

\subsection{Activation Functions}
\label{sec:anns:an:act_functions}

An \index{activation function}activation function is used to model the action
potential of the \index{AN}\ac{AN} \cite{ref:ziv:1994, ref:hodgkin:1952}, and is
a function over the augmented \index{net input signal}net input signal. It is
the mechanism that models whether an \index{AN}\ac{AN} fires or not. When the
output produced by the activation function surpasses some threshold value
$\theta$, we consider that neuron to have fired an output signal.
\index{activation function}Activation functions model \textit{phase shift}. In
the context of classification problems, these \index{activation
function}activation functions form decision boundaries between classes. In the
context of regression problems, these \index{activation function}activation
functions try to approximate some function that maps the input data to some
target data.

In general, activation functions produce a non-linear mapping of
$\mathbb{R}^{I}$ to the range $[0,1]$ or $[-1,1]$ as shown below:

\begin{equation}
    f_{AN}: \mathbb{R}^{I} \rightarrow [0,1],
    \label{eq:an_function_mapping_0_1}
\end{equation}

\noindent or

\begin{equation}
    f_{AN}: \mathbb{R}^{I} \rightarrow [-1,1].
    \label{eq:an_function_mapping_minus_1_1}
\end{equation}

\noindent Research in the past has presented many \index{activation
function}activation functions \cite{ref:karlik:2011} of varying ranges. However,
this dissertation focuses on the \ac{ReLU} \cite{ref:jarrett:2009, ref:nair:2010}, the
\ac{LReLU} \cite{ref:maas:2013}, the \index{sigmoid}sigmoid \cite{ref:lecun:1988} and the
\index{hyperbolic tangent}hyperbolic tangent \cite{ref:lin:2008} \index{activation
function}activation functions. These \index{activation function}activation
functions are presented next.

\subsubsection{Rectified Linear Unit}
\label{sec:anns:an:act_functions:relu}

The \ac{ReLU} \index{activation function}activation function is an
\index{activation function}activation function defined in the positive part of
its argument. It is defined as:

@@

\noindent An advantage of \ac{ReLU} is that it is not susceptible to the
\index{vanishing gradient problem}vanishing gradients problem \cite{ref:xu:2015,
ref:maksutov:2018} which occurs when gradients in the first layers of a
multi-layer \ac{ANN} approach zero, and have no effect on the training process.

\subsubsection{Leaky Rectified Linear Unit}
\label{sec:anns:an:act_functions:leaky_relu}

>>>>

\begin{quotation}
    \noindent ``Men ought to know that from the brain, and from the brain only, arise our pleasures, joy, laughter and jests, as well as our sorrows, pains, griefs, and tears.''
\end{quotation}
\begin{flushright}
    Hippocrates
\end{flushright}

\noindent
The human brain is one of the most complicated biological organs in nature. It gives us the ability to think and learn. The field of \ac{ML} has taken a lot of inspiration from the biological brain which lead to the development of \acp{ANN} \cite{ref:rosenblatt:1958}. \acp{ANN} are used throughout this dissertation as the \textit{model} to be trained using \acp{HH}.

The purpose of this chapter is to provide the necessary background information needed on \acp{ANN} and is thus structured into the following sections.

@@

    \item \textbf{Section \ref{sec:anns:an}} introduces the \ac{AN}. Brief discussions follow on input, \index{weights}weights, \index{biases}biases, \index{net input signal}net input signal, \index{activation functions}activation functions, and output.

    \item \textbf{Section \ref{sec:anns:ann}} introduces the \ac{ANN}. Brief discussions follow on \ac{ANN} architecture, topology, and \acp{FFNN}.

    \item \textbf{Section \ref{sec:anns:training}} provides details on the training process, training sets, \index {supervised learning}supervised learning, \index{error function}error functions, performance measures and stopping conditions.

    \item \textbf{Section \ref{sec:anns:summary}} provides a brief summary of the chapter.
\end{itemize}

\index{Biological Neuron}
\section{Biological Neuron}\label{sec:anns:bn}

This section introduces the \ac{BN} and provides the necessary background information that show how the \ac{BN} has inspired the development of the \ac{AN}.

@@

\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.95\textwidth]{biological_neuron.pdf}
    \caption[The \index{biological neuron}biological Neuron]{An illustration of the \index{biological neuron}biological Neuron.}
    \label{fig:biological_neuron}
\end{figure}

\noindent
The main components of the \ac{BN} include the cell body, \index{dendrites}\textit{dendrites} and the \index{axon}\textit{axon}. \Acp{NN} are formed through the connections between the \index{axons}axons and \index{dendrites}dendrites of various neurons. This is known as \index{synaptogenesis}\textit{synaptogenesis} \cite{ref:huttenlocher:1997}. Such a connection is referred to as a \index{synapse}\textit{synapse}. Communication takes place, through the synapse, by electro-chemical pulse and is often referred to as an \textit{activation} or \textit{action potential}.  Communication signals propagate from the \index{dendrites}dendrites, through the cell body to the axon of a neuron, provoking a signal in the post-synaptic neuron \cite{ref:engelbrecht:2007}. The greater the connection between two neurons, the stronger the communication.  Kennedy \cite{ref:kennedy:2016} defines stronger \index{synapse}synapses as ones that contribute more depolarisation to the neural membrane upon activation than weaker ones. Stronger \index{Synapse}synapses have a higher probability of generating an action potential in the post-synaptic neuron. During activation, the pre-synaptic neuron release neurotransmitters that bind to the post-synaptic neuron \cite{ref:khanacademy:synapse}. The frequent release of these molecules cause the \index{synapse}synapse to grow. Connections that grow over time yield stronger signals (learning), while connections that are weak propagate low intensity signals and vanish over time (forgetting). The ability of synapses to strengthen and weaken over time is known as \index{synaptic plasticity}\textit{synaptic plasticity} \cite{ref:huttenlocher:1997}.

\index{Artificial Neuron}
\section{Artificial Neuron}\label{sec:anns:an}

This section introduces the \ac{AN}. Brief discussion follow on the various components that make up the \ac{AN}. These components include inputs, weights, biases, \index{net input signal}net input signal, and output of the \ac{AN}.

\subsection{Components}\label{sec:anns:components}

The \ac{AN} implements a nonlinear mapping from $\mathbb{R}^{I}$ to
$\mathbb{R}^{T}$, usually in the ranges $[0,1]$ or $[-1,1]$, depending on the
activation function used \cite{ref:engelbrecht:2007} and is given
in Equation \eqref{eq:an_function_mapping} below.

@@

\noindent
In Equation \eqref{eq:an_function_mapping} above, $f_{AN}$ is the mapping function produced by the \ac{AN}, $I$ is the total number of dimensions of the input in real-number space ($\mathbb{R}$), and $T$ is the total number of dimensions of the target (desired output) in real-number space.

The \ac{AN} implements various components and is illustrated in Figure \ref{fig:artificial_neuron}.

\begin{figure}[htpb]
    \centering
    \includegraphics[width=\textwidth]{artificial_neuron.pdf}
    \caption[The \index{artificial neuron}artificial neuron]{An illustration of the \index{artificial neuron}artificial neuron.}
    \label{fig:artificial_neuron}
\end{figure}

\noindent
Each of the components of the \ac{AN} is inspired by some element of the \ac{BN}. Brief descriptions of these are given as follows.

\begin{itemize}
    \item \textbf{Input:} The input models the activations received from the pre-synaptic neuron or from some environment sensor. Input is represented by the input vector $\vec{x}$, in Figure \ref{fig:artificial_neuron}.

    \item \textbf{Weights:} The weights model the \index{synapse}synapses and connection strengths. Weights are represented by the weight vector $\vec{v}$, in Figure \ref{fig:artificial_neuron}.

    \item \index{net input signal}\textbf{Net input signal:} The \index{net input signal}net input signal models the net resulting activations from all connected pre-synaptic neurons or environment sensors. The \index{net input signal}net input signal is represented by $net$, in Figure \ref{fig:artificial_neuron}.

    \item \textbf{Biases:} The biases model a mechanism introduced to influence the strength of the activation (output signal) of the \index{AN}\ac{AN}. The bias is represented by $\theta$, in Figure \ref{fig:artificial_neuron}.

    \item \index{activation function}\textbf{Activation function:} The activation function models the action potential of the \ac{BN}\index{BN} based on the net input signal. The \index{activation function}activation function is represented by $f$, in Figure \ref{fig:artificial_neuron}.

    \item \textbf{Output:} The output models the activation by the post-synaptic neuron. Output is represented by the output vector $\vec{y}$, in Figure \ref{fig:artificial_neuron}.
\end{itemize}

\noindent
The following sections provide a detailed discussion on each of the above mentioned components.

\subsection{Input}\label{sec:anns:an:input}

Input signals are $I$-dimensional vectors of numerical values that are obtained either through some environment sensor or from other \acp{AN}. Input signals are often referred to as \index{features}\textit{features} or \index{independent variables}\textit{independent variables} \cite{ref:francis:2001}. Throughout this dissertation, a single input vector is referred to as a pattern. Input data must first be pre-processed before it is presented to the \ac{AN}. Input pre-processing techniques are presented in the following sections.

\subsubsection{Input Pre-Processing}\label{sec:anns:an:input:input_pre_processing}

Input pre-processing improves the training process and contributes to the success of the practical application of the \ac{ANN} \cite{ref:kuzniar:2017}. The primary purpose of input pre-processing is to modify the input data so that it can better match predicted output data. There are many techniques to consider when pre-processing input data. These techniques includes methods to encode data to certain formats, to scale to the correct ranges and to deal with incomplete, invalid or irrelevant data. For the purposes of this dissertation, only \textit{data encoding} and \textit{normalisation} is considered.

\index{Encoding}
\subsubsection{Encoding}\label{sec:anns:an:input:encoding}

For qualitative data such as categories, the class labels have to be converted from textual to numerical representations. In general, class labels are encoded as numerical vectors \cite{ref:srinidhi:2018, ref:brownlee:2017:one-hot}. One such encoding technique is referred to as \index{one-hot encoding}\textit{one-hot encoding}. Harris and Harris \cite{ref:harris:2010} describe \index{One-Hot Encoding}one-hot encoding as a group of vector elements where the classes are represented with a single high element (represented by $1$) and all the others low (represented by $0$). Each element $y_k$, where $k$ is the $k$-th element of the encoded vector $\vec{y}$, uniquely refers to a class. An activation, represented by a value of $1$ at a given index thus represents a class associated with that index.

\index{Normalisation}
\subsubsection{Normalisation}\label{sec:anns:an:input:normalisation}

Numerical data must be normalised and scaled appropriately for the \ac{AN}. In general, input data is normalised to a range that is appropriate for the \index{activation function}activation function being used by the \ac{AN}. Normalisation and scaling yields input data that is comparable to the output produced by the \ac{AN}. For the purposes of this disseration, \index{min-max scaler}min-max scaler \cite{ref:al:2006} and \index{standard score scaler}standard score scaler \cite{ref:jain:2005} are considered.

\index{Min-Max Scalar}
\subsubsection{Min-Max Scaler}\label{sec:anns:an:input:min_max_scaler}

The \index{min-max scaler}min-max scaler, also called \index{unity-based normalisation}\textit{unity-based normalisation}, scales values to the range $[0,1]$. The \index{min-max scaler}min-max scaler is used as a pre-processing technique for target values when the \index{sigmoid}\textit{sigmoid} \index{activation function} activation function is used. The \index{min-max scaler}min-max scaler is given in Equation \eqref{eq:min_max_scaler} below.

@@

\noindent
In Equation \eqref{eq:min_max_scaler} above, $x_{i,p}^{'}$ is the normalised form of $x_{i,p}$, the $i$-th dimension of the input vector $\vec{x}_p$, $p \in \{1,2, \dots, P \}$, where $P$ is the total number of input patterns, $x_{i_{min}}$ and $x_{i_{max}}$ are respectfully the minimum and maximum values of the $i$-th dimension for all input vectors $\vec{x}_p$.

\index{Standard Score Scalar}
\subsubsection{Standard Score Scaler}\label{sec:anns:an:input:standard_score_scaler}

The \index{standard score scaler}standard score scaler, also known as the \index{z-score scaler}\textit{z-score scaler}, scales values by subtracting the mean and scaling to the unit variance of each dimension $i$ for all input patterns $\vec{x}_p, p \in \{1,2, \dots, P\}$. The \index{standard score scaler}standard score scaler is used as a pre-processing technique for target values when the \index{hyperbolic tangent}\textit{hyperbolic tangent} \index{activation function}activation function\index{activation function} is used. The \index{standard score scaler}standard score scaler is given in Equation \eqref{eq:standard_score_scaler} below.

@@

\noindent
In Equation \eqref{eq:standard_score_scaler} above, $\mu_i$ and $\sigma^2_i$ are respectfully the mean and unit variance of the $i$-th dimension of all input vectors $\vec{x}_p$.

\subsection{Weights}\label{sec:anns:an:weights}

The connection strength that synapses in the \acs{BN} have is modeled in the \ac{AN} as weight vectors of numerical values associated with each dimension of the input. Weights can either dampen or strengthen the input by some negative or positive numerical value. Changes in the weight associated with a feature changes the influence that that particular feature has on the predicted output. Finding the correct value for each weight, such that the \ac{AN} yields optimal output, is an optimisation problem \cite{ref:thierens:1993}. Research has shown that weight initialisation plays an important role in the efficient training of \acp{ANN} \cite{ref:thimm:1995}.

\index{Weight Initialisation}
\subsubsection{Weight Initialisation}\label{sec:anns:an:weights:initialisation}

Weight initialisation is the process by which candidate solutions (represented by the weights of the \ac{AN}) to the problem are ``placed'' in the search-space. Weight initialisation influences the speed of convergence, the probability of convergence and the generalisation capabilities of \acp{ANN} \cite{ref:fernandez:2001}. Finding the optimal initialisation values for weights is non-trivial and can be seen as another optimisation problem \cite{ref:de:2016, ref:erdogmus:2003, ref:yam:2000}.

\index{Weight Initialisation}Weight initialisation is dependent on the \index{activation function}activation function used. If weights are initialised as values that are too small, the vanishing gradients problem can occur \cite{ref:hanin:2018}. If weights are initialised as values that are too big, output of the activation function would not be in the active range. This leads to unit saturation, and exploding gradients can occur \cite{ref:hanin:2018, ref:yadav:2018}.

Many different weight initialisation techniques have been developed \cite{ref:erdogmus:2003}. Some techniques have been developed to address the problems that are mentioned above \cite{ref:yadav:2018}. For the purposes of this dissertation, focus is put on \index{random uniform sampling}random uniform sampling, \index{Glorot uniform sampling}\textit{Glorot uniform (Xavier)} sampling and \index{Glorot normal sampling}\textit{Glorot normal} sampling \cite{ref:glorot:2010}. Brief discussions on each of these weight initialisation techniques are presented as follows.

\index{random uniform sampling}
\subsubsection{Random Uniform Sampling}\label{sec:anns:an:weights:random_uniform_sampling}

\index{random uniform sampling}Random uniform sampling initialises weights uniformly in the range $[\omega_{min}, \omega_{max}]$, written as $w_{i} \sim \textit{U} (\omega_{min}, \omega_{max})$, where the $\omega_{min}$ and $\omega_{max}$ are respectfully the lower and upper bounds of the uniform distribution. Suggested parameter values are $(-1, 1)$ or $(-0.5, 0.5)$ \cite{ref:nguyen:1990}.

\index{Glorot uniform sampling}
\subsubsection{Glorot Uniform Sampling}\label{sec:anns:an:weights:glorot_uniform_sampling}

@@

\index{Glorot Normal Sampling}
\subsubsection{Glorot Normal Sampling}\label{sec:anns:an:weights:glorot_normal_sampling}

@@

\index{net input signal}
\subsection{Net Input Signal}\label{sec:anns:an:net_input}

\acp{AN} accumulate the net resulting input signal from all input dimensions into a value called the \index{net input signal}\textit{net input signal}, expressed as $net$. This signal is passed to the \index{activation function}activation function, which provokes an artificial action potential in the \ac{AN}.

A common way by which the \index{net input signal}net input signal is calculated is by means of \acp{SU} \cite{ref:engelbrecht:2007}, which compute the \index{net input signal}net input signal as the weighted sum of all input signals as shown in Equation \eqref{eq:summation_units} below.

@@

\noindent
In Equation \eqref{eq:summation_units} above, $x_{i}$ is the $i$-th dimension of the input vector $\vec{x}$, and $v_{i}$ is the $i$-th dimension of the weight vector $\vec{v}$, associated with the given input dimension.

\subsection{Biases}\label{sec:anns:an:biases}

A bias/threshold term $\theta$ is introduced to help translate the output of the \index{activation function}activation function \cite{ref:benitez:1997}. The value of $\theta$ can be learned during the training process along with the weights of the \ac{ANN}. In order to simplify equations, the input and weight vectors are augmented such that the input and hidden layers have an additional neuron/unit, called the \textit{bias unit} \cite{ref:engelbrecht:2007}. The net input signal can then be rewritten to consider the bias unit, leading to an augmented net input signal.

\subsubsection{Augmented Net Input Signal}\label{sec:anns:an:biases:augmented_net_input_signal}

The augmented net input signal, that includes the bias unit, has the form $net^{'} = net - \theta$. The value of $\theta$ is learned along with the weights such that $\theta = x_{i+1}v_{i+1}$. A constant value $x_{i+1} = -1$ can be used. The net input signal as given in Equation \eqref{eq:summation_units} then changes as is shown in Equation \eqref{eq:augmented_vectors} below.

@@

\subsection{Activation Functions}\label{sec:anns:an:act_functions}

An \index{activation function}activation function is used to model the action potential of the \index{AN}\ac{AN} \cite{ref:ziv:1994, ref:hodgkin:1952}. The \index{activation function}activation function takes as a parameter, the augmented \index{net input signal}net input signal. When the output produced by the activation function surpasses some threshold value $\tau$, we consider that neuron to have ``fired'' an output signal. \index{activation function}Activation functions thus model \textit{phase shift}. In the context of classification problems, \index{activation function}activation functions form decision boundaries between classes. In the context of regression problems, \index{activation function}activation functions try to approximate some function that maps the input data to some target data.

In general, activation functions produce a non-linear mapping of $\mathbb{R}^{I}$ to the range $[0,1]$ or $[-1,1]$ as shown in Equations \eqref{eq:an_function_mapping_0_1} and \eqref{eq:an_function_mapping_minus_1_1} below.

\begin{equation}
    f_{AN}: \mathbb{R}^{I} \rightarrow [0,1]
    \label{eq:an_function_mapping_0_1}
\end{equation}

\begin{equation}
    f_{AN}: \mathbb{R}^{I} \rightarrow [-1,1]
    \label{eq:an_function_mapping_minus_1_1}
\end{equation}

\noindent
Many different \index{activation function}activation functions have been developed \cite{ref:karlik:2011}. For the purposes of this dissertation, focus is put on the \acf{ReLU} \cite{ref:jarrett:2009, ref:nair:2010}, the \acf{LReLU} \cite{ref:maas:2013}, the \index{sigmoid}sigmoid \cite{ref:lecun:1988} and the \index{hyperbolic tangent}hyperbolic tangent \cite{ref:lin:2008} \index{activation function}activation functions.

\subsubsection{Rectified Linear Unit}\label{sec:anns:an:act_functions:relu}

The \acf{ReLU} \index{activation function}activation function is an \index{activation function}activation function defined in the positive part of its argument and is given in Equation \eqref{eq:relu} below.

@@

\noindent
An advantage of \acs{ReLU} is that it is not susceptible to the \index{vanishing gradient problem}vanishing gradients problem \cite{ref:xu:2015, ref:maksutov:2018} which occurs when gradients in the first layers of a multi-layer \acs{ANN} approach zero, and have no effect on the training process.

\subsubsection{Leaky Rectified Linear Unit}\label{sec:anns:an:act_functions:leaky_relu}
